### 2016-07-15 cross_compilation.tex
Not updated commits

\subsection{Toolchain Setup}
To compile in a given architecture that is aimed for a different one,
the testbed administrator needs to install a toolchain in its \ac{PC}.
The toolchain is mandatory due to the different processor architectures
where the source can be compiled from / to.

Given that compiling a toolchaing can be an ardous task, we get the toolchain
recommended for the \ac{ARM} architecture of the \ac{Raspi}s.
This toolchain is available from \cite{2016rpitoolchain} and
it already contains the binaries for different compilers based on
\texttt{gcc 4.9}. In what follows, we extract the binaries adjusting
them to our coding style and compiling convention. For this, we use the
\texttt{\$TOOLCHAIN} directory as the working directory. However,
the testbed administrator may choose some other working directory of
its preference if desired. First, we create the toolchain directory:

@@

This helps the \ac{OS} to recognize the location of the compiler
command everytime a new shell is opened. The \texttt{.profile} should now
contain the lines we inserted. There might be more other
code in your file. Leave other parts unmodified.
% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$HOME/.profile}@>
<@\textcolor{gray}{
---------------------------}
\Reactivatenumber @>
...
export TOOLCHAINDIR="$HOME/toolchains"
export TOOLCHAINBINARY="raspberry-gxx49-arm-g++"
PATH="$PATH:${TOOLCHAINDIR}/arm-rpi-4.9.3-linux-gnueabihf/bin"
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

>>>>

\subsection{Toolchain Setup}
To compile in a given architecture that is aimed for a different one,
the testbed administrator needs to install a toolchain in its \ac{PC}.
The toolchain is mandatory due to the different processor architectures
where the source can be compiled from.
Given that compiling a toolchaing can be an ardous task, we get the toolchain
recommended for the \ac{ARM} architecture of the \ac{Raspi}s.
This toolchain is available from \cite{2016rpitoolchain} and
it already contains the binaries for different compilers based on
\texttt{gcc 4.9}. In what follows, we extract the binaries adjusting
them to our coding style and compiling convention. For this, we use the
\texttt{\$TOOLCHAIN} directory as the working directory. However,
the testbed administrator may choose some other working directory of
its preference if desired. First, we create the toolchain directory:

@@

This helps the \ac{OS} to recognize the location of the compiler
command everytime a new shell is opened. The \texttt{.profile} should now
contain the lines we inserted. There might be other
code in the file of other testbed administrators. We recommend to leave
other parts unmodified.
% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$HOME/.profile}@>
<@\textcolor{gray}{
---------------------------}
\Reactivatenumber @>
...
export TOOLCHAINDIR="$HOME/toolchains"
export TOOLCHAINBINARY="raspberry-gxx49-arm-g++"
PATH="$PATH:${TOOLCHAINDIR}/arm-rpi-4.9.3-linux-gnueabihf/bin"
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

### 2016-07-15 overlay_fs.tex
HTTP ongoing work
Not updated commits

\label{sec:ovelay_fs}
In principle, our procedure modifies the image file only once in
the testbed administrator \ac{PC} when its setup is made. Also, keeping
this image in the \ac{Raspi}s provides the same initial system for all
the devices. This system is later distributed locally to a memory card
in each device. However, if we do not make any further modifications during
the image setup, any files created after the initial boot of a \ac{Raspi}
will remain in the memory card. This is cumbersome to maintain since
the size of the memory card is relatively reduced (8 GB) and there
might be various users utilizing the testbed. Also, different testbed
users could be interested in running their experiments in a fresh
rebooted system with the original customized image. Nevertheless, we
emphasize that this step is not necessary if the tesbed administrator
wants to consider only persistent storage for its devices. A use case
for this scenario could be a single user for the testbed or when a testbed
administrator only wants to setup a few \ac{Raspi}s.

Therefor, if both persistent and non-persistent storage are required for
the \ac{Raspi}s, here we present the steps to setup an overlay filesystem.
This type of filesystem enables an \textit{upper} filesystem to overlay
into a \textit{lower} filesystem. Whenever a file is requested, the upper
filesystem will forward the request to the lower filesystem in case it
does not have it itself. If the upper filesystem has the requested file,
it will simpy return the file. This idea can be used in our setup to mount
the root filesystem (i.e. Raspbian lite) in the \ac{Raspi}s during startup
as read-only filesystem. Thus, on the one hand, the image configuration files
will remain after a reboot and will not be modifiable. On the other hand,
to enable the possibility of non-persistent changes, we overlay the upper
filesystem that is mounted in the \ac{Raspi} \ac{RAM}, i.e. \texttt{/tmp}
as read/writable on top of the lower root filesystem. Thus, reading a file
may return a file from the lower filesytem, but if it is stored, it will
be saved in the upper filesystem. Accessing this file again will return
the stored file from the upper layer. After a reboot, all the stored files
in the upper filesystem will be erasured.

\subsection{Overlaying Filesystem Installation}

Assuming that we are still using the \texttt{chroot} commands into the
Raspbian lite root filesystem for installing packages, we can setup the
overlay filesystem at this point of the procedure. There already exists
implementations overlaying the root filesystem. We use an implementation
available at the Git repository in \cite{overlayroot}. Since we have
installed git in a previous step, we clone the repository. The
command block below stores it in \texttt{/tmp} which is really mounted in
\ac{RAM}. Then, all the files stored here will disappear a \ac{Raspi} is
rebooted.

@@

Although these commands might be output some warnings, they should
successfully generate working initial \ac{RAM} filesystems. Later, an
initial \ac{RAM} filesystem will need to be called by the bootloader.
In Raspbian this is done by adding a command to \texttt{config.txt} file
in the boot partition. If the system should be run in a \ac{Raspi} version 1,
then use \texttt{init.gz}, otherwise use \texttt{init-v7.gz} as follows
\todot{What happens if I have both a Raspi 1, 2 or 3 in my testbed? Do I have
to have different images after this point?}:

@@

If the steps are successfully executed up to this point, the customized image
is available in the \texttt{\$\{IMAGE\}.img} file and ready to be deployed
into the \ac{Raspi}s. In the following section, we indicate how to proceed
with the writing of the image into various memory cards, although a more
complex setup with a \ac{HTTP} server is also presented in
Section~\ref{sec:http} of the Appendix.

@@

It is \textit{very
important} to write to the correct device as everything will be overwritten.
To avoid removing information from the wrong device, a testbed administrator
can insert / eject a memory card a couple of times and using the
\texttt{sudo lsblk} and \texttt commands to deduce the correct device.
For our case, the device was \texttt{/dev/sdb} in an Ubuntu distribution and
\texttt{/dev/mmcblk0} in an Archlinux distribution. Once identified, to write
the image to a memory card, the following command is used:

>>>>

\label{sec:overlay_fs}
In principle, our procedure modifies the image file only once in
the testbed administrator \ac{PC} when its setup is made. Also, keeping
this image in the \ac{Raspi}s provides the same initial system for all
the devices. If we do not make any further modifications during
the image setup, any files created after the initial boot of a \ac{Raspi}
will remain in the memory card. This is cumbersome to maintain since
the size of the memory card is relatively reduced (8 GB) and there
might be various users utilizing the testbed. Also, different testbed
users could be interested in running their experiments in a fresh
rebooted system with the original customized image. Nevertheless, we
emphasize that this step is not necessary if the tesbed administrator
wants to consider only persistent storage for its devices. A use case
for this scenario could be a single user for the testbed or when a testbed
administrator only wants to setup a few \ac{Raspi}s.

If both persistent and non-persistent storage are required for
the \ac{Raspi}s, here we present the steps to setup an overlay filesystem.
This type of filesystem enables an \textit{upper} filesystem to overlay
into a \textit{lower} filesystem. Whenever a file is requested, the upper
filesystem will forward the request to the lower filesystem in case it
does not have it itself. If the upper filesystem has the requested file,
it will simpy return the file. This idea can be used in our setup to mount
the root filesystem (i.e. Raspbian Lite) in the \ac{Raspi}s during startup
as read-only filesystem. Thus, on the one hand, the image configuration files
will remain after a reboot and will not be modifiable. On the other hand,
to enable the possibility of non-persistent changes, we overlay the upper
filesystem that is mounted in the \ac{Raspi} \ac{RAM}, i.e. \texttt{/tmp}
as read / writable on top of the lower root filesystem. Thus, reading a file
may return a file from the lower filesytem, but if it is stored, it will
be saved in the upper filesystem. Accessing this file again will return
the stored file from the upper layer. After a reboot, all the stored files
in the upper filesystem will be retrieved.

\subsection{Filesystem Installation}

Assuming that we are still in the \texttt{chroot} environment into the
Raspbian Lite root filesystem for installing packages, we can setup the
overlay filesystem at this point of the procedure. There already exists
implementations overlaying the root filesystem. We use an implementation
available at the Git repository in \cite{overlayroot}. Since we have
installed \texttt{git} in a previous step, we clone the repository. The
command block below stores it in \texttt{/tmp} which is really mounted in
\ac{RAM}. Then, all the files stored here will disappear a \ac{Raspi} is
rebooted.

@@

Although these commands might output some warnings, they should
successfully generate working initial \ac{RAM} filesystems. Later, an
initial \ac{RAM} filesystem will need to be called by the bootloader.
In Raspbian this is done by adding a command to \texttt{config.txt} file
in the boot partition. If the system should be run in a \ac{Raspi} version 1,
then use \texttt{init.gz}, otherwise use \texttt{init-v7.gz} as follows
\todot{What happens if I have both a Raspi 1, 2 or 3 in my testbed? Do I have
to have different images after this point?}:

@@

Again, if the \texttt{----recursive} option is not available, the
procedure in Section~\ref{sec:umount} of the Appendix could be followed.
If the steps are successfully executed up to this point, the customized image
is available in the \texttt{\$\{IMAGE\}.img} file and ready to be deployed
into the \ac{Raspi}s. In the following section, we indicate how to proceed
with the writing of the image into various memory cards, although a more
complex setup with a \ac{HTTP} server is also presented in
Section~\ref{sec:http} of the Appendix.

@@

It is \textit{very important} to write to the correct device as
everything will be overwritten. To avoid removing information from the
wrong device, a testbed administrator can insert and eject a memory
card a couple of times and using the \texttt{sudo lsblk} and \texttt{df -h}
commands to deduce the correct device. For our case, the device was
\texttt{/dev/sdb} in an Ubuntu distribution and
\texttt{/dev/mmcblk0} in an Archlinux distribuion. Once identified, to write
the image to a memory card, the following command is used:

### 2016-07-15 testbed_http.tex
Not updated commits

\label{sec:testbed_http}
Creating manual copies of the customized image for each memory card can
be also a time-consuming task that is prone to errors and become
cumbersome. Therefore, we present an optional more advanced \ac{HTTP}
image fetching setup for the existing image to be downloaded from a
\ac{HTTP} server. We consider this an optional step since a testbed
administrator might opt for not including this as part of his / her setup.
We start our description consideri

At the end of Section~\ref{sec:overlay_fs}, the final customized image was
proposed to be splitted into three partitions. One for the boot files, one
containing the root file system and one for the home directory. This is made
considering an overlay filesystem approach. In this section, instead of
writing the image direclty to the memory cards, we present how the
partitions can be extracted from the image, and how the boot partition can
be configured to download the root filesystem and home directory partitions
from a \ac{HTTP} server during startup. The advantage of this setup is that
we completely automate the image fetch which significantly eases its
maintainance since any changes are required to be made only in the \ac{HTTP}
server image. We start by declaring the variables for the path names of each
partition:

@@

Then, we setup a loopback device of the existing image and use \texttt{dd} to
extract the boot and root partitions from the image:
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
# dd if=${DEV}p1 of=${HTTPBOOTIMAGE}.img bs=4M
# dd if=${DEV}p2 of=${HTTPROOTIMAGE}.img bs=4M
# losetup -d ${DEV}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now it is time to configure the boot- and root-image just extracted before
the content of the boot image is copied to the a memory card and the root
image is uploaded to a \ac{HTTP} server.

@@

\subsubsection{Prepare sd-card}

@@

\subsubsection{(Optional) Compressing root image}

@@

\subsubsection{Try it out}

>>>>

\label{sec:testbed_http}
Creating manual copies of the customized image for each memory card can
be also a time-consuming task that is prone to errors and become
cumbersome. Therefore, we present an optional more advanced \ac{HTTP}
image fetching setup for the existing image to be downloaded from a
\ac{HTTP} server. We consider this an optional step since a testbed
administrator might opt for not including this as part of his / her setup.

At the end of Section~\ref{sec:overlay_fs}, the final customized image was
proposed to be splitted into three partitions. One for the boot files, one
containing the root file system and one for the home directory. This is made
considering an overlay filesystem approach. In this section, instead of
writing the image directly to the memory cards, we present how the
partitions can be extracted from the image, and how the boot partition can
be configured to download the root filesystem and home directory partitions
from a \ac{HTTP} server during startup. The advantage of this setup is that
we completely automate the image fetch which significantly eases its
maintainance since any changes are required to be made only in the \ac{HTTP}
server image. We start by declaring the variables for the path names of each
partition:

@@

We set loopback devices for each of the image partitions and use \texttt{dd} to
extract them from the image:
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
# dd if=${DEV}p1 of=${HTTPBOOTIMAGE}.img bs=4M
# dd if=${DEV}p2 of=${HTTPROOTIMAGE}.img bs=4M
# dd if=${DEV}p3 of=${HTTPHOMEIMAGE}.img bs=4M
# losetup -d ${DEV}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, we configure each of these partitions. The content of the boot image is
copied and modified just once into a memory card and the root
and home images is uploaded to a \ac{HTTP} server. Even though we still
have to manually configure and set the memory cards, we only have
to make this one time, since the \ac{Raspi}s will fetch the root filesystem
from the \ac{HTTP} server every they are turned on.

@@

\subsection{Prepare sd-card}

@@

\subsection{Compressing root image (Optional)}

@@

\subsection{Try it out}
